Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?

### 2020-06-21 docs/00-manuscripts/ieee/ieee-manuscript.tex
address v4 review from Alex

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
	This study analyzed highly-correlated, feature-rich datasets from hyperspectral remote sensing data using multiple machine and statistical-learning methods.
	The effect of filter-based feature-selection methods on predictive performance was compared.
	Also, the effect of multiple expert-based and data-driven feature sets, derived from the reflectance data, was investigated.
	Feature wise, defoliation at trees (\%) was modeled as a function of band reflectance.
	Variable importance was assessed using permutation-based feature importance.

	With respect to the used models, support vector machine (SVM) outperformed others such as random forest (RF), extreme gradient boosting (XGBoost), lasso (L1) and ridge (L2) regression by at least three percentage points.
	The combination of certain feature sets showed small increases in predictive performance while no substantial differences between single feature sets were observed.
	Filter methods helped in certain dataset-learner-filter settings to achieve better predictive performances while ensemble filters did not have a substantial impact on performance.
	For certain dataset-learner combinations, filter methods helped to improve predictive performance and lower computational runtime.

@@

	The dataset used in this work was suboptimal with respect to the achieved predictive performances and more training data is needed for more generalizable conclusions.
	Filter methods show potential to be helpful in high-dimensional situations while still keeping the ability to interpret feature effects of fitted models, which is an essential constrain in environmental modeling studies.

@@

	\acro{AGB}{above-ground biomass}
	\acro{ALE}{accumulated local effects}
	\acro{ALS}{airborn laser scanning}
	\acro{ANN}{artificial neural network}
	\acro{AUROC}{area under the receiver operating characteristics curve}
	\acro{BRT}{boosted regression trees}
	\acro{CART}{classification and regression trees}
	\acro{CNN}{convolutional neural networks}
	\acro{CV}{cross-validation}
	\acro{DAP}{digital aerial photogrammetry}
	\acro{ENM}{environmental niche modeling}
	\acro{FPR}{false positive rate}
	\acro{FFS}{forward feature selection}
	\acro{FS}{feature selection}
	\acro{GAM}{generalized additive model}
	\acro{GBM}{gradient boosting machine}
	\acro{GLM}{generalized linear model}
	\acro{ICGC}{Institut Cartografic I Geologic de Catalunya}
	\acro{IQR}{interquartile range}
	\acro{MARS}{multivariate adaptive regression splines}
	\acro{MBO}{model-based optimization}
	\acro{MEM}{maximum entropy model}
	\acro{ML}{machine learning}
	\acro{NDII}{normalized difference infrared index}
	\acro{NIR}{near-infrared}
	\acro{NRI}{normalized ratio index}
	\acro{OLS}{ordinary least squares}
	\acro{LiDAR}{light detection and ranging}
	\acro{LOWESS}{locally weighted scatter plot smoothing}
	\acro{PISR}{potential incoming solar radiation}
	\acro{PCA}{principal component analysis}
	\acro{PDP}{partial dependence plots}
	\acro{PLS}{partial least-squares}
	\acro{RBF}{radial basis function}
	\acro{RF}{random forest}
	\acro{RMSE}{root mean square error}
	\acro{RR}{ridge regression}
	\acro{RSS}{residual sum of squares}
	\acro{SAR}{synthetic aperture radar}
	\acro{SDM}{species distribution modeling}
	\acro{SMBO}{sequential-based model optimization}
	\acro{SVM}{support vector machine}
	\acro{TPR}{true positive rate}
	\acro{VI}{vegetation index}
	\acro{XGBoost}{extreme gradient boosting}
\end{acronym}
\renewcommand{\IEEEiedlistdecl}{\relax}% reset back

@@

\IEEEPARstart{T}{he} use of \ac{ML} algorithms for analyzing remote sensing data has seen a huge increase in the last decade \cite{lary2016}.
This goes in line with the increased availability of remote sensing imagery, especially since the launch of the first Sentinel satellite in the year 2014.
At the same time, the implementation and usability of learning algorithms has been greatly simplified with many contributions from the open-source community.
Scientists can nowadays relatively easily process large amounts of (environmental) information using various learning algorithms.
This makes it possible to extend the benchmark comparison matrix of studies in a semi-automated way, possibly stumbling across unexpected findings of process settings that would never have been tested otherwise \cite{ma2015}.

@@

Machine-learning methods in combination with remote sensing data are used in many environmental fields such as vegetation cover analysis or forest carbon storage mapping \cite{mascaro2014, urban2018}.
The ability of predicting into unknown space qualifies these tools as a promising toolset for such tasks.
One aspect of this research field is to enhance the understanding of biotic and abiotic triggers, for example by analyzing defoliation at trees \cite{hawrylo2018}.

Other approaches for analyzing forest health include temporal change detection \cite{zhang2016} or describing the current health status of forests on a stand level \cite{townsend2012}.
In such studies, the defoliation of trees serves as a proxy for forest health by describing the impact of biotic and abiotic pest triggers \cite{townsend2012, goodbody2018}.

Vegetation indices have shown the potential to provide valuable information when analyzing forest health \cite{jiang2014, adamczyk2015}.
Most vegetation indices were developed with the aim of being sensitive to changes of specific wavelength regions, serving as a proxy for underlying plant processes. 
However, often enough indices developed for different purposes than the one to be analyzed can help to explain complex relationships.
This emphasizes the need to extract as much information as possible from the available input data to generate promising features which can help to understand the modeled relationship.
A less known index type which can be derived from spectral information is the \ac{NRI}.
In contrast to most vegetation indices, \ac{NRI}s do not use an expert-based formula following environmental heuristics but instead makes use of a data-driven feature engineering approach by combining (arbitrary) combinations of spectral bands.
Especially when working with hyperspectral data, thousands of \ac{NRI} features can be derived this way.

@@

Despite its popularity in environmental modeling, there are no studies so far which used machine-learning algorithms in combination with remote sensing data to analyze defoliation on a tree level.
This study aims to close this gap by analyzing defoliation at trees in northern Spain using airborne hyperspectral data.
The methodology of this study uses machine-learning methods in combination with feature selection and hyperparameter tuning.
In addition, feature importance and feature effects are evaluated.
Incorporating the idea of creating data-driven \ac{NRI}s, this study also discusses the practical problems of high-dimensionality in environmental modeling \cite{trunk1979, xu2016}.

% mention why FS is used / important
Even though \ac{ML} algorithms are capable of handling highly-correlated input variables, model fitting becomes computationally more demanding, and model interpretation more complex.
Feature selection approaches can help to address this issue, reducing possible in the feature space, simplify model interpretability and possibly enhance predictive performance \cite{cai2018}.

% clarify the focus of this study
This study shows how high-dimensional datasets can be handled effectively with machine-learning methods while still being conduct inference on the fitted models.
The predictive power of non-linear methods and their ability to handle highly-correlated predictors is combined with common and new approaches for assessing feature importance and feature effects.
However, this study clearly focuses on investigating the effects of filter methods and feature set types on predictive performance rather than on interpreting feature effects.

That said, the research questions of this study are the following:

@@

	\item Do different (environmental) feature sets show differences in performance when modeling defoliation at trees?

@@

	\item How are feature-selection methods influencing the predictive performance of the models?

@@

Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain.
The trees in the plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum}, leading to a spread of cankers or defoliation \cite{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation at trees (serving as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} which ranges from 0 - 100 (in \%) (\autoref{fig:defol-distr}).
It is assumed that the fungi infect the trees through open wounds, possibly caused by previous hail damage \cite{iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \cite{ganley2009}.

@@

The \textit{Pinus radiata} plots of this study, namely Laukiz1, Laukiz2, Luiando and Oiartzun, are located in the northern part of the Basque Country (\autoref{fig:study_area}).
Oiartzun has the most observations (n = 559) while Laukiz2 shows the largest area size (1.44 ha).
All plots besides Luiando are located nearby the coast (\autoref{fig:study_area}).
In total 1808 observations are available Laukiz1 = 559, Laukiz2 = 451, Luiando = 301, Oiartzun = 497).
The data was surveyed in September 2016.

@@

The airborne hyperspectral data was acquired during two flight campaigns on September 28th and October 5th 2016, both around noon.
The images were taken by an AISAEAGLE-II sensor.
All preprocessing steps (geometric, radiometric, atmospheric) were conducted by the \ac{ICGC}.
The first four bands were corrupted, leaving 122 bands with valid information.
Additional metadata information is available in \autoref{tab:hyperparameter_limits}.

@@

\begin{table}[t]
\centering
\caption[t]{Specifications of hyperspectral data.}
\begingroup
\begin{tabular}{ll}
	\\
	Characteristic         & Value                               \\
	\toprule
	Geometric resolution   & 1 m                                 \\
	Radiometric resolution & 12 bit                              \\
	Spectral resolution    & 126 bands (404.08 nm --- 996.31 nm) \\
	Correction:            & Radiometric, geometric, atmospheric
\end{tabular}
\endgroup\label{tab:hyperparameter_limits}
\end{table}

@@

To account for geometric offsets within the hyperspectral data, which were reported with up to 1 m from \ac{ICGC}, a buffer of two meters around the centroid of each tree was used when extracting the reflectance values.
A pixel was considered to fall into a tree's buffer zone if the centroid of the respective pixel was touched by the buffer.
All of those pixels formed the final reflectance value of a single tree and were used as the base information to derive all additional feature sets.
In total, \(\frac{121*122}{2} = 7471\) NRIs were calculated due to the first four bands of the sensor being corrupted.

@@

The case of a feature-rich dataset comes with several challenges for both model fitting and evaluation.

\begin{itemize}
	\item Model fitting times increase.
	\item Noise is possibly introduced into models by highly-correlated variables \cite{johnstoneiainm.2009}.
	\item Model interpretation and prediction become more challenging \cite{johnstoneiainm.2009}.
\end{itemize}

@@

% Filter methods
The concept of filters originates from the idea of ranking features using certain heuristics of an algorithm \cite{guyon2003}.
Some filter methods can only deal with specific types of variables (numeric or nominal).
Filters only rank features, they do not decide which covariates to drop or keep \cite{drotar2015}.
The selection which features to keep for model fitting is usually done within the optimization phase of the model fitting, along with the hyperparameter tuning.
Essentially, the number of covariates in the model is treated as a hyperparameter of the model.
The goal is to optimize the number of features, after ranking was done, to the point at which the model achieves the best performance.

% Ensemble filter methods
Besides the concept of choosing a specific filter method to rank variables, studies showed that combining several filters using statistical operations such as 'minimum' or 'mean' are able to enhance the predictive performance of the resulting models, especially when applied to multiple datasets \cite{abeel2010, drotar2017a}.
This approach is referred to as 'ensemble filtering' \cite{dietterich2000}.
Ensemble filters align with the recent rise of the 'ensemble' approach in machine learning which uses the idea of stacking to combine the predictions of multiple models, aiming to enhance predictive performance \cite{polikar2012, feurer2015, bolon-canedo2019}.
In this work the 'Borda' ensemble filter was applied \cite{drotar2017a}.
It's final feature order is determined by the sum of all single filters ranks.

@@

where \(H\) is the conditional entropy of the response variable (class) or the feature (attribute), respectively.
In order to use this method with a numeric response (percentage defoliation at trees), the variable was discretized into equal bins and treated as a class variable.
After feature rank correlations of $> 0.9$ betwforeen different bin sizes were observed in a side analysis, \texttt{\(n_{bin}\) = 10} was found to be a reasonable setting to go with.

@@

Random forest and {SVM} are well established algorithms widely used in (environmental) modeling.
Extreme Gradient Boosting (commonly abbreviated as XGBoost) has shown promising results in benchmark studies in recent years.
Penalized regression is a statistical modeling technique capable of dealing with highly-correlated covariates by penalizing the coefficients of the model \cite{hastie2001}.
Common penalties are 'lasso' (L1) and 'ridge' (L2).
Ridge does not remove variables from the model (penalization to zero) but just shrinks them to effectively zero, keeping them in the model.
The combination of both penalties is called 'elastic net' but was not used in this work.

@@

Three feature sets were used in this study with each representing a different way of feature engineering:

@@

Even though the feature-selection step should be solely left to the filter methods in this study,it was ensured a priori to account for features with a pairwise correlation of 1.
Having such features within the data can cause undesired effects during model fitting and feature importance calculation.
Hence, after having calculated all pair-wise correlations between features, for pairs which exceeded the threshold of 0.9999999999, the feature with the largest mean absolute correlation across all variables was removed from the dataset.
This process was repeated $p$ times, each time calculating a fresh correlation matrix.

This preprocessing step reduced the amount of covariates to 122 (HR), 86 (VI) and 7467 (NRI).

@@

In this work, an initial design of 30 randomly composed hyperparameter settings in combination with a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is the substantial reduction of the tuning budget which is required to find a setting close to the global optimization minimum.
\ac{MBO} may outperform methods that do not use information from previous iterations, such as random search or grid search \cite{bergstra2012}.

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage (\cite{binder2020}).
For \ac{PCA}, the number of principal components was tuned instead.
The RF hyperparameter \texttt{\(m_{try}\)} was modified into power transformed \(p^{m_{try}}\) (ranging from 1 to $p$, with \(p\) being the number of features of the dataset) to work on relative values of the respective dataset's feature count.
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features left after optimizing the feature percentage during tuning.

@@

In total the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods + 2 $\times$ 6 L1 and L2 models).

@@

Estimating feature importance for datasets with highly-correlated features is a complicated task.
The correlation between covariates makes it challenging to calculate an unbiased estimate for single features \cite{molnar2019}.
Methods like \ac{PDP} or permutation-based approaches may produce unreliable estimates in such scenarios because unrealistic situations between covariates are created \cite{molnar2019}.
Estimating feature importance for machine learning methods is a complicated procedure for which many different approaches, model-specific and agnostic, exist \cite{friedman2001, hastie2001, greenwell2018}.
The development of robust methods which enable an unbiased estimation of feature importance for highly-correlated variables are subject to current research.

@@

The algorithm implementations of the following packages have been used: xgboost \cite{chen2016} (\textit{Extreme Gradient Boosting}), kernlab \cite{kernlab} (Support Vector Machine) and glmnet \cite{glmnet} (penalized regression).
The filter implementations of the following packages have been used: praznik \cite{praznik}, FSelectorRcpp \cite{fselectorrcpp}.
Package mlr \cite{mlr} was used for all modeling related steps.
drake \cite{drake} was used for structuring the work and reproducibility.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).
Besides the availability of code and manuscript sources, a static webpage is available at (\url{https://github.com/pat-s/2019-feature-selection}),listing more side-analyses that were carried out during the creation of this study.

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
Overall, the response variable \enquote{tree defoliation} could be modeled with an \ac{RMSE} of 28 \% percentage points (p.p.).
SVM showed no differences in RMSE across feature sets whereas other learners (RF, SVM, XGBoost, lasso and ridge) differed up to seven percentage points (\autoref{fig:perf-result}).
Ridge faced major issues in four tasks due to one observation which was predicted off the response scale (i.e. $> 100$).
SVM showed the best overall performance with a difference of around three percentage points to the next best model (RF) (\autoref{tab:best-learner-perf}).
Performance differences between test folds were large: Predicting on Luiando resulted in an RMSE of 9.0 for learner SVM (without filter) but up to 54.26 p.p. when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).

@@

% filter vs. no filter
% - Substantial effect for XGBoost and RF
% - SVM unaffected
When comparing the usage of filters against using no filter at all, there was only one instance (XGBoost on the NRI task) when a model without filtering scored a better performance than the best filtered one (\autoref{fig:filter-effects-no-filter}).
For SVM, all filters and \enquote{no filter} achieved the same performance on tasks VI and NRI even though \autoref{fig:perf-result} lists "No Filter" as the best option.

@@

% Selected percentage of features across plots
High differences were observed between the amount of features selected during tuning for the subsequent fitting process.
Most features were selected during optimization for plot Laukiz2 and least for Laukiz1 (\autoref{tab:tune-perc-sel-features}).
RF used only one feature for plots Luiando and Oiartzun (0.00004 \% of 7675) while for Laukiz1 34 and for Laukiz2 230 features were seen as optimal during optimization.
In contrast, XGBoost and SVM used in all cases but Laukiz1 (less than 50 features) more than two-third of all available features.

@@

% plot performance results
\begin{figure} [t!]
	\centering
	\begin{center}
		\includegraphics[width=0.48\textwidth] {performance-results-1.pdf}
		\caption{Predictive performance (RMSE) of models across tasks. Suffix 'CV' denotes that the learner was optimized using internal 10-fold CV while prefix 'MBO' means that Bayesian optimization was used for hyperparameter optimization. Abbreviations on the vertical axis refer to the combinations of feature sets on which each model was scored on. Labels represent the feature selection method (NF = no filter, Car = 'Carscore', Info = 'Information Gain', Borda = 'Borda'). The second value of each label shows the RMSE value of the respective setting.}\label{fig:perf-result}
	\end{center}
\end{figure}

@@

The ALE plots show a small relative change compared to their respective mean effects for all chosen features ($< \mid 0.001 \mid$) (\autoref{fig:fi-hr-ale}).
Most ALE curves show a high dynamic across the value range of each variable, within their respective effect range.

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
	This study analyzed highly-correlated, feature-rich datasets from hyperspectral remote sensing data using multiple machine and statistical-learning methods.
	The effect of filter-based feature-selection methods on predictive performance was compared.
	Also, the effect of multiple expert-based and data-driven feature sets, derived from the reflectance data, was investigated.
	Defoliation of trees (\%) was modeled as a function of reflectance, and variable importance was assessed using permutation-based feature importance.

	Overall support vector machine (SVM) outperformed others such as random forest (RF), extreme gradient boosting (XGBoost), lasso (L1) and ridge (L2) regression by at least three percentage points.
	The combination of certain feature sets showed small increases in predictive performance while no substantial differences between individual feature sets were observed.
	For some combinations of learners and feature sets, filter methods achieved better predictive performances than the unfiltered feature sets, while ensemble filters did not have a substantial impact on performance.

@@

	More training data and replication in similar benchmarking studies is needed for more generalizable conclusions.
	Filter methods have the potential to be helpful in high-dimensional situations and are able to improve the interpretation of feature effects in fitted models, which is an essential constraint in environmental modeling studies.

@@

	\acro{AGB}{above-ground biomass}
	\acro{ALE}{accumulated local effects}
	\acro{ALS}{airborne laser scanning}
	\acro{ANN}{artificial neural network}
	\acro{AUROC}{area under the receiver operating characteristics curve}
	\acro{BRT}{boosted regression trees}
	\acro{CART}{classification and regression trees}
	\acro{CNN}{convolutional neural networks}
	\acro{CV}{cross-validation}
	\acro{DAP}{digital aerial photogrammetry}
	\acro{ENM}{environmental niche modeling}
	\acro{FPR}{false positive rate}
	\acro{FFS}{forward feature selection}
	\acro{FS}{feature selection}
	\acro{GAM}{generalized additive model}
	\acro{GBM}{gradient boosting machine}
	\acro{GLM}{generalized linear model}
	\acro{ICGC}{Institut Cartografic i Geologic de Catalunya}
	\acro{IQR}{interquartile range}
	\acro{MARS}{multivariate adaptive regression splines}
	\acro{MBO}{model-based optimization}
	\acro{MEM}{maximum entropy model}
	\acro{ML}{machine learning}
	\acro{NDII}{normalized difference infrared index}
	\acro{NIR}{near-infrared}
	\acro{NRI}{normalized ratio index}
	\acro{OLS}{ordinary least squares}
	\acro{LiDAR}{light detection and ranging}
	\acro{LOWESS}{locally weighted scatter plot smoothing}
	\acro{PISR}{potential incoming solar radiation}
	\acro{PCA}{principal component analysis}
	\acro{PDP}{partial dependence plots}
	\acro{PLS}{partial least-squares}
	\acro{RBF}{radial basis function}
	\acro{RF}{random forest}
	\acro{RMSE}{root mean square error}
	\acro{RR}{ridge regression}
	\acro{RSS}{residual sum of squares}
	\acro{SAR}{synthetic aperture radar}
	\acro{SDM}{species distribution modeling}
	\acro{SMBO}{sequential-based model optimization}
	\acro{SVM}{support vector machine}
	\acro{TPR}{true positive rate}
	\acro{VI}{vegetation index}
	\acro{XGBoost}{extreme gradient boosting}
\end{acronym}
\renewcommand{\IEEEiedlistdecl}{\relax}% reset back

@@

\IEEEPARstart{T}{he} use of \ac{ML} algorithms for analyzing remote sensing data has seen a huge increase in the last decade \cite{lary2016}.
This goes in line with the increased availability of remote sensing imagery, especially since the launch of the first Sentinel satellite in the year 2014.
At the same time, the implementation and usability of learning algorithms has been greatly simplified with many contributions from the open-source community.
Scientists can nowadays relatively easily process large amounts of (environmental) information using various learning algorithms.
This makes it possible to extend the benchmark comparison matrix of studies in a semi-automated way, possibly stumbling across unexpected findings of process settings that would never have been explored otherwise \cite{ma2015}.

@@

Machine-learning methods in combination with remote sensing data are used in many environmental fields such as vegetation cover analysis or forest carbon storage mapping \cite{mascaro2014, urban2018}.
The ability of predicting into unknown space qualifies these tools as a promising toolset for such tasks.
One aspect of this research field is to enhance the understanding of biotic and abiotic triggers, for example by analyzing tree defoliation \cite{hawrylo2018}.

Other approaches for analyzing forest health include change detection \cite{zhang2016} or describing the current health status of forests on a stand level \cite{townsend2012}.
In such studies, the defoliation of trees serves as a proxy for forest health by describing the impact of biotic and abiotic pest triggers \cite{townsend2012, goodbody2018}.

Vegetation indices have shown the potential to provide valuable information when analyzing forest health \cite{jiang2014, adamczyk2015}.
Most vegetation indices were developed with the aim of being sensitive to changes of specific wavelength regions, serving as a proxy for underlying plant processes.
However, often enough indices developed for different purposes than the one to be analyzed can help to explain complex relationships.
This emphasizes the need to extract as much information as possible from the available input data to generate promising features which can help to understand the modeled relationship.
A less known index type which can be derived from spectral information is the \ac{NRI}.
In contrast to most vegetation indices, \ac{NRI}s do not use an expert-based formula following environmental heuristics but instead makes use of a data-driven feature engineering approach by combining (all possible) combinations of spectral bands.
Especially when working with hyperspectral data, thousands of \ac{NRI} features can be derived this way.

@@

Despite their popularity in environmental modeling, there are no studies so far which used machine-learning algorithms in combination with remote sensing data to analyze defoliation at the tree level.
This study aims to close this gap by analyzing tree defoliation in northern Spain using airborne hyperspectral data.
The methodology of this study uses machine-learning methods in combination with feature selection and hyperparameter tuning.
In addition, feature importance and feature effects are evaluated.
Incorporating the idea of creating data-driven \ac{NRI}s, this study also discusses the practical problems of high dimensionality in environmental modeling \cite{trunk1979, xu2016}.

% mention why FS is used / important
Even though \ac{ML} algorithms are capable of handling highly correlated input variables, model fitting becomes computationally more demanding, and model interpretation more complex.
Feature selection approaches can help to address this issue, reducing possible noise in the feature space, simplify model interpretability and possibly enhance predictive performance \cite{cai2018}.

% clarify the focus of this study
This study shows how high-dimensional datasets can be handled effectively with machine-learning methods while still being able to interpret the fitted models.
The predictive power of non-linear methods and their ability to handle highly correlated predictors is combined with common and new approaches for assessing feature importance and feature effects.
However, this study clearly focuses on investigating the effects of filter methods and feature set types on predictive performance rather than on interpreting feature effects.

Considering these opportunities and challenges, the research questions of this study are the following:

@@

	\item Do different (environmental) feature sets show differences in performance when modeling defoliation of trees?

@@

	\item How do different feature selection methods influence the predictive performance of the models?

@@

Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain.
The trees in the plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum}, leading to a spread of cankers or defoliation \cite{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation of trees (serving as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} which ranges from 0 - 100 (in \%) (\autoref{fig:defol-distr}).
It is assumed that the fungi infect the trees through open wounds, possibly caused by previous hail damage \cite{iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \cite{ganley2009}.

@@

The \textit{Pinus radiata} plots of this study, namely Laukiz1, Laukiz2, Luiando and Oiartzun, are located in the northern part of the Basque Country (\autoref{fig:study_area}).
Oiartzun has the most observations (n = 559) while Laukiz2 shows the largest area size (1.44 ha).
All plots besides Luiando are located within 100 km from the coast (\autoref{fig:study_area}).
In total 1808 observations are available Laukiz1 = 559, Laukiz2 = 451, Luiando = 301, Oiartzun = 497).
Field surveys were conducted in September 2016.

@@

The airborne hyperspectral data was acquired during two flight campaigns on September 28th and October 5th 2016, both around noon.
The images were taken by an AISAEAGLE-II sensor.
All preprocessing steps (geometric, radiometric, atmospheric) were conducted by the \ac{ICGC}.
The first four bands were corrupted, leaving 122 bands with valid information.
Additional metadata information is available in \autoref{tab:hyperparameter_metadata}.

@@

\begin{table}[t]
\centering
\caption[t]{Specifications of hyperspectral data.}
\begingroup
\begin{tabular}{ll}
	\\
	Characteristic         & Value                               \\
	\toprule
	Geometric resolution   & 1 m                                 \\
	Radiometric resolution & 12 bit                              \\
	Spectral resolution    & 126 bands (404.08 nm --- 996.31 nm) \\
	Correction:            & Radiometric, geometric, atmospheric
\end{tabular}
\endgroup\label{tab:hyperparameter_metadata}
\end{table}

@@

To account for geometric offsets within the hyperspectral data, which were reported with up to 1 m from \ac{ICGC}, a buffer of two meters around the centroid of each tree was used when extracting the reflectance values.
A pixel was considered to fall into a tree's buffer zone if the centroid of the respective pixel was touched by the buffer.
All of those pixels formed the final reflectance value of a single tree and were used as the base information to derive all additional feature sets.
In total, \(\frac{121*122}{2} = 7471\) NRIs were calculated.

@@

High-dimensional, feature-rich datasets come with with several challenges for both model fitting and evaluation.

\begin{itemize}
	\item Model fitting times increase.
	\item Noise is possibly introduced into models by highly correlated variables \cite{johnstoneiainm.2009}.
	\item Model interpretation and prediction become more challenging \cite{johnstoneiainm.2009}.
\end{itemize}

@@

% Filter methods
The concept of filters originates from the idea of ranking features following a score calculated by an algorithm \cite{guyon2003}.
Some filter methods can only deal with specific types of variables (numeric or nominal).
Filters only rank features, they do not decide which covariates to drop or keep \cite{drotar2015}.
The selection which features to keep for model fitting is usually done within the optimization phase of the model fitting, along with the hyperparameter tuning.
Essentially, the number of covariates in the model is treated as a additional hyperparameter of the model.
The goal is to optimize the number of ranked features to the point at which the model achieves the best performance.

% Ensemble filter methods
Besides the concept of choosing a specific filter method to rank variables, studies showed that combining several filters using statistical operations such as 'minimum' or 'mean' is able to enhance the predictive performance of the resulting models, especially when applied to multiple datasets \cite{abeel2010, drotar2017a}.
This approach is referred to as 'ensemble filtering' \cite{dietterich2000}.
Ensemble filters align with the recent rise of the 'ensemble' approach in machine learning which uses the idea of stacking to combine the predictions of multiple models, aiming to enhance predictive performance \cite{polikar2012, feurer2015, bolon-canedo2019}.
In this work the 'Borda' ensemble filter was applied \cite{drotar2017a}.
Its final feature order is determined by the sum of all single filters ranks.

@@

where \(H\) is the conditional entropy of the response variable (class) or the feature (attribute), respectively.
In order to use this method with a numeric response (percentage defoliation of trees), the variable was discretized into equal bins and treated as a class variable.
After feature rank correlations of $> 0.9$ between different bin sizes were observed in a side analysis, \texttt{\(n_{bin}\) = 10} was found to be a reasonable setting to go with.

@@

Random forest and {SVM} are well established algorithms widely used in (environmental) remote sensing.
Extreme Gradient Boosting (commonly abbreviated as XGBoost) has shown promising results in benchmark studies in recent years.
Penalized regression is a statistical modeling technique capable of dealing with highly-correlated covariates by penalizing the coefficients of the model \cite{hastie2001}.
Common penalties are 'lasso' (L1) and 'ridge' (L2).
Ridge does not remove variables from the model (penalization to zero) but just shrinks them to effectively zero, keeping them in the model.

@@

Three feature sets were used in this study, each representing a different approach to feature engineering:

@@

Even though the feature-selection step should be solely left to the filter methods in this study, it was ensured a priori to account for features with a pairwise correlation of 1.
Having such features within the data can cause undesired effects during model fitting and feature importance calculation.
Hence, after having calculated all pair-wise correlations between features, for pairs which exceeded the threshold of $1 - 10^{-10}$, the feature with the largest mean absolute correlation across all variables was removed from the dataset.

This preprocessing step reduced the number of covariates to 122 (HR), 86 (VI) and 7467 (NRI).

@@

In this work, an initial design of 30 randomly composed hyperparameter settings in combination with a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is a substantial reduction of the tuning budget that is required to find a setting close to the global optimization minimum.
\ac{MBO} may outperform methods that do not use information from previous iterations, such as random search or grid search \cite{bergstra2012}.

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage (\cite{binder2020}).
For \ac{PCA}, the number of principal components was tuned instead.
The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_try = p^{x_try}$.
The exponent $x_try$, (ranging from 1 to $p_sel$, where $p_sel$ is the number of features selected with the corresponding filter settings), was optimized between 0 (one randomly selected feature) and 0.5 ($\sqrt{p_sel}$ randomly selected candidate features).
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features left after optimizing the feature percentage during tuning.

@@

In total the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods and for the L1/L2 models, 6 feature sets $\times$ 2 models.

@@

Estimating feature importance for datasets with highly correlated features is a complicated task for which many different approaches, model-specific and agnostic, exist \cite{friedman2001, hastie2001, greenwell2018}..
The correlation between covariates makes it challenging to calculate an unbiased estimate for single features \cite{molnar2019}.
Methods like \ac{PDP} or permutation-based approaches may produce unreliable estimates in such scenarios because unrealistic situations between covariates are created \cite{molnar2019}.
The development of robust methods which enable an unbiased estimation of feature importance for highly correlated variables are subject to current research.

@@

The algorithm implementations of the following packages have been used: xgboost \cite{chen2016} (\textit{Extreme Gradient Boosting}), kernlab \cite{kernlab} (Support Vector Machine) and glmnet \cite{glmnet} (penalized regression).
The filter implementations of the following packages have been used: praznik \cite{praznik}, FSelectorRcpp \cite{fselectorrcpp}.
Package mlr \cite{mlr} was used for all modeling related steps.
drake \cite{drake} was used for structuring the work and reproducibility.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).
Besides the availability of code and manuscript sources, a static webpage is available at (\url{https://github.com/pat-s/2019-feature-selection}), listing more side-analyses that were carried out during the creation of this study.

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
Overall, the response variable \enquote{tree defoliation} could be modeled with an \ac{RMSE} of 28 percentage points (p.p.).
SVM showed no differences in RMSE across feature sets whereas other learners (RF, SVM, XGBoost, lasso and ridge) differed up to seven percentage points (\autoref{fig:perf-result}).
Ridge faced major issues in four tasks due to one observation which was predicted off the response scale (i.e. $> 100$).
SVM showed the best overall performance with a mean difference of around three percentage points to the next best model (RF) (\autoref{tab:best-learner-perf}).
Performance differences between test folds were large: Predicting on Luiando resulted in an RMSE of 9.0 p.p. for learner SVM (without filter) but up to 54.3 p.p. when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).

@@

% filter vs. no filter
% - Substantial effect for XGBoost and RF
% - SVM unaffected
When comparing the usage of filters against using no filter at all, there was only one instance (XGBoost on the NRI task) where a model without filtering scored a better performance than the best filtered one (\autoref{fig:filter-effects-no-filter}).
For SVM, all filters and \enquote{no filter} achieved the same performance on tasks VI and NRI even though \autoref{fig:perf-result} lists "No Filter" as the best option.

@@

% Selected percentage of features across plots
% FIXME: Lösung einfügen nach Klärung mit Alex
Large differences were observed between the numbers of features selected during tuning for the subsequent fitting process.
Most features were selected during optimization for plot Laukiz2 and least for Laukiz1 (\autoref{tab:tune-perc-sel-features}).
RF used only one feature for plots Luiando and Oiartzun (0.00004 \% of 7675) while for Laukiz1 34 and for Laukiz2 230 features were seen as optimal during optimization.
In contrast, XGBoost and SVM used in all cases but Laukiz1 (less than 50 features) more than two-thirds of all available features.

@@

% plot performance results
\begin{figure} [t!]
	\centering
	\begin{center}
		\includegraphics[width=0.48\textwidth] {performance-results-1.pdf}
		\caption{Predictive performance (RMSE) of models across tasks. Suffix 'CV' denotes that the learner was optimized using internal 10-fold CV while prefix 'MBO' means that model-based optimization was used for hyperparameter optimization. Abbreviations on the vertical axis refer to the combinations of feature sets on which each model was scored on. Labels represent the feature selection method (NF = no filter, Car = 'Carscore', Info = 'Information Gain', Borda = 'Borda'). The second value of each label shows the RMSE value of the respective setting.}\label{fig:perf-result}
	\end{center}
\end{figure}

@@

Most ALE plots show a small absolute change in ALE compared to their respective mean effects (= 0) for all chosen features (less than $\pm 0.001$) (\autoref{fig:fi-hr-ale}).
Also, a medium-to-high relative dynamic over the respective reflectance value range of each variable was observed.
